Lesson 4
========================================================

***

### Scatterplots and Perceived Audience Size
Notes:

***

### Scatterplots
Notes:

```{r Scatterplots}
setwd('~/r/nanodegree/explor_data_analys/')
library(ggplot2)
pf <- read.csv('OLD/pseudo_facebook.tsv', sep = '\t')

qplot(x = age, y = friend_count, data = pf)
qplot(age, friend_count, data = pf)

ggplot(aes(x = age, y = friend_count), data = pf) + 
  geom_point()

summary(pf$age)

ggplot(aes(x = age, y = friend_count), data = pf) + 
  geom_point() + xlim(13, 90)

```

***

#### What are some things that you notice right away?
Response:

***

### ggplot Syntax
Notes:

```{r ggplot Syntax}

```

***

### Overplotting
Notes:

```{r Overplotting}
ggplot(aes(x = age, y = friend_count), data = pf) + 
  geom_point(alpha = 1/20) + xlim(13, 90)

ggplot(aes(x = age, y = friend_count), data = pf) + 
  geom_jitter(alpha = 1/20) + xlim(13, 90)
```

#### What do you notice in the plot?
Response:

***

### Coord_trans()
Notes:

```{r Coord_trans()}
ggplot(aes(x = age, y = friend_count), data = pf) + 
  geom_point(alpha = 1/20) + xlim(13, 90) + coord_trans(y = "sqrt")

ggplot(aes(x = age, y = friend_count), data = pf) + 
  geom_point(alpha = 1/20, position = position_jitter(h = 0)) + 
  xlim(13, 90) + coord_trans(y = "sqrt")
```

#### Look up the documentation for coord_trans() and add a layer to the plot that transforms friend_count using the square root function. Create your plot!

```{r}

```

#### What do you notice?

***

### Alpha and Jitter
Notes:

```{r Alpha and Jitter}
ggplot(aes(x = age, y = friendships_initiated), data = pf) + 
  geom_jitter(alpha = 1/20) + xlim(13, 90)
```

***

### Overplotting and Domain Knowledge
Notes:

***

### Conditional Means
Notes:

```{r Conditional Means}
#install.packages('dplyr')
library(dplyr)

age_groups <- group_by(pf, age)
pf.fc_by_age <- summarise(age_groups, 
          friend_count_mean = mean(friend_count),
          friend_count_median = median(friend_count),
          n = n())

pf.fc_by_age <- arrange(pf.fc_by_age, age)

head(pf.fc_by_age)

pf.fc_by_age <- pf %>% 
  group_by(age) %>%
  summarise(age_groups, 
          friend_count_mean = mean(friend_count),
          friend_count_median = median(friend_count),
          n = n()) %>%
  arrange(pf.fc_by_age)

head(pf.fc_by_age)

head(pf.fc_by_age, 20)

```

Create your plot!

```{r Conditional Means Plot}
ggplot(aes(x = age, y = friend_count_mean), data = pf.fc_by_age) + 
  geom_line()
```

***

### Overlaying Summaries with Raw Data
Notes:

```{r Overlaying Summaries with Raw Data}
ggplot(aes(x = age, y = friend_count), data = pf) +
  xlim(13, 90) +
  geom_point(alpha = 0.05,
             position = position_jitter(h = 0),
             color = 'orange') +
  coord_trans(y = 'sqrt') +
  geom_line(stat = 'summary', fun.y = mean) +
  geom_line(stat = 'summary', fun.y = quantile, probs = 0.1,
            linetype = 3, color = 'blue') +
  geom_line(stat = 'summary', fun.y = quantile, probs = 0.5,
            color = 'blue') +
  geom_line(stat = 'summary', fun.y = quantile, probs = 0.9,
            linetype = 2, color = 'blue')

ggplot(aes(x = age, y = friend_count), data = pf) +
  coord_cartesian(xlim = c(13, 70), ylim = c(0, 1000)) +
  geom_point(alpha = 0.05,
             position = position_jitter(h = 0),
             color = 'orange') +
  geom_line(stat = 'summary', fun.y = mean) +
  geom_line(stat = 'summary', fun.y = quantile, probs = 0.1,
            linetype = 3, color = 'blue') +
  geom_line(stat = 'summary', fun.y = quantile, probs = 0.5,
            color = 'blue') +
  geom_line(stat = 'summary', fun.y = quantile, probs = 0.9,
            linetype = 2, color = 'blue')
```

#### What are some of your observations of the plot?
Response:

***

### Moira: Histogram Summary and Scatterplot
See the Instructor Notes of this video to download Moira's paper on perceived audience size and to see the final plot.

Notes:

***

### Correlation
Notes:

```{r Correlation}
cor(pf$age, pf$friend_count)

cor.test(pf$age, pf$friend_count, method = 'pearson')

with(pf, cor.test(age, friend_count, method = 'pearson'))
```

Look up the documentation for the cor.test function.

What's the correlation between age and friend count? Round to three decimal places.
Response:

***

### Correlation on Subsets
Notes:

```{r Correlation on Subsets}
with(subset(pf, age <= 70), cor.test(age, friend_count))

with(subset(pf, age <= 70), cor.test(age, friend_count,
                                     method = 'spearman'))
```

***

### Correlation Methods
Notes:

***

## Create Scatterplots
Notes:

```{r}
ggplot(aes(x = www_likes_received, y = likes_received), data = pf) + geom_point() + coord_trans(y = "sqrt")

```

***

### Strong Correlations
Notes:

```{r Strong Correlations}
ggplot(aes(x = www_likes_received, y = likes_received), data = pf) +
  geom_point() +
  xlim(0, quantile(pf$www_likes_received, 0.95)) +
  ylim(0, quantile(pf$likes_received, 0.95)) +
  geom_smooth(method = 'lm', color = 'red')
```

What's the correlation betwen the two variables? Include the top 5% of values for the variable in the calculation and round to 3 decimal places.

```{r Correlation Calcuation}
cor(pf$www_likes_received, pf$likes_received)
```

Response:

***

### Moira on Correlation
Notes:

***

### More Caution with Correlation
Notes:

```{r More Caution With Correlation}
#install.packages('alr3')
library(alr3)

data(Mitchell)
?Mitchell
```

Create your plot!

```{r Temp vs Month}
ggplot(Mitchell, aes(x = Temp, y = Month)) +
  geom_point()

cor(Mitchell$Temp, Mitchell$Month)
```

***

### Noisy Scatterplots
a. Take a guess for the correlation coefficient for the scatterplot.

b. What is the actual correlation of the two variables?
(Round to the thousandths place)

```{r Noisy Scatterplots}

```

***

### Making Sense of Data
Notes:

```{r Making Sense of Data}
ggplot(Mitchell, aes(x = Month, y = Temp)) +
  geom_point() +
  scale_x_discrete(breaks = seq(0, max(Mitchell$Month), 12))

ggplot(aes(x=(Month%%12),y=Temp),data=Mitchell)+ 
  geom_point()

dcor.ttest() # in the energy package implements a non-parametric test of the independence of two variables. The test correctly rejects the independence.



```

***

### A New Perspective

What do you notice?
Response:

Watch the solution video and check out the Instructor Notes!
Notes:

***

### Understanding Noise: Age to Age Months
Notes:

```{r Understanding Noise: Age to Age Months}

ggplot(aes(x = age, y = friend_count_mean), data = pf.fc_by_age) +
  geom_line()

head(pf.fc_by_age, 10)
pf.fc_by_age[17:19, ]
```

***

### Age with Months Means

```{r Age with Months Means}
pf$age_with_months <- (2013 - pf$dob_year) + ((12 - pf$dob_month)/12.0)


```

Programming Assignment
```{r Programming Assignment}
library(dplyr)
age_month_groups <- group_by(pf, age_with_months)

#rm(pf.fc_by_age_months)

pf.fc_by_age_months <- pf %>%
  group_by(age_with_months) %>%
  summarise(friend_count_mean = mean(friend_count),
            friend_count_median = median(friend_count),
            n = n()) %>%
  arrange(age_with_months)

head(pf.fc_by_age_months)

ggplot(pf.fc_by_age_months, aes(x=friend_count_mean, y=age_with_months)) +
  geom_point()

ggplot(pf.fc_by_age_months, aes(x=age_with_months, y=friend_count_mean)) +
  coord_cartesian(xlim =c(0,70)) +
  geom_line()

ggplot(aes(x=age_with_months, y=friend_count_mean),
       data = subset(pf.fc_by_age_months, age_with_months < 71)) +
  geom_line()

```

***

### Noise in Conditional Means

```{r Noise in Conditional Means}

```

***

### Smoothing Conditional Means
Notes:

```{r Smoothing Conditional Means}
p1 <- ggplot(pf.fc_by_age_months, aes(x=age_with_months, y=friend_count_mean)) +
  coord_cartesian(xlim =c(0,70)) +
  geom_line()

p2 <- ggplot(aes(x=age_with_months, y=friend_count_mean),
       data = subset(pf.fc_by_age_months, age_with_months < 71)) +
  geom_line()

p3 <- ggplot(aes(x = round(age / 5) * 5, y = friend_count),
       data = subset(pf, age < 71)) +
  geom_line(stat = 'summary', fun.y = mean)

library(gridExtra)
grid.arrange(p2, p1, p3, ncol = 1)

p1 <- ggplot(pf.fc_by_age_months, aes(x=age_with_months, y=friend_count_mean)) +
  coord_cartesian(xlim =c(0,70)) +
  geom_line() +
  geom_smooth()

p2 <- ggplot(aes(x=age_with_months, y=friend_count_mean),
       data = subset(pf.fc_by_age_months, age_with_months < 71)) +
  geom_line() +
  geom_smooth()

p3 <- ggplot(aes(x = round(age / 5) * 5, y = friend_count),
       data = subset(pf, age < 71)) +
  geom_line(stat = 'summary', fun.y = mean)

grid.arrange(p2, p1, p3, ncol = 1)
```

***

### Which Plot to Choose?
Notes:

***

### Analyzing Two Variables
Reflection:

***

# PROBLEM SET
```{r}
data(diamonds)
names(diamonds)
ggplot(diamonds, aes(x=price, y=x)) +
  geom_point()

cor(diamonds$price, diamonds$x)
cor(diamonds$price, diamonds$y)
cor(diamonds$price, diamonds$z)

ggplot(diamonds, aes(x=price, y=depth)) +
  geom_point(alpha=1/100) +
  scale_x_continuous(breaks = seq(0, max(diamonds$price), 2))

cor(diamonds$price, diamonds$depth)
cor(diamonds$depth, diamonds$price)

ggplot(diamonds, aes(x=carat, y=price)) +
  geom_point() +
  xlim(0,quantile(diamonds$carat, 0.99)) +
  ylim(0,quantile(diamonds$price, 0.99))

diamonds$volume <- (diamonds$x*diamonds$y*diamonds$z)

ggplot(diamonds, aes(x=price, y=volume)) +
  geom_point()

#library(plyr)
count(diamonds$volume == 0)
count(diamonds$volume < 800)

detach("package:plyr", unload=TRUE)

diamonds$volume != 0 

diamonds$volume < 800

these <- subset(diamonds, volume != 0 & volume < 800)

these

cor(these$price, these$volume)

ggplot(these, aes(x=price, y=volume)) +
  geom_point() +
  geom_smooth()

pf.fc_by_age_months <- pf %>%
  group_by(age_with_months) %>%
  summarise(friend_count_mean = mean(friend_count),
            friend_count_median = median(friend_count),
            n = n()) %>%
  arrange(age_with_months)

head(pf.fc_by_age_months)

#library(plyr)
library(dplyr)
#rm(diamondsByClarity)
clarity_groups <- group_by(diamonds, clarity) %>%
  arrange(clarity)

diamondsByClarity <- clarity_groups %>%
  summarise(mean_price = mean(as.numeric(price)),
            median_price = median(as.numeric(price)),
            min_price = min(as.numeric(price)),
            max_price = max(as.numeric(price)),
            n = n())

head(diamondsByClarity)


diamonds_by_clarity <- group_by(diamonds, clarity)
diamonds_mp_by_clarity <- summarise(diamonds_by_clarity, mean_price = mean(price))

diamonds_by_color <- group_by(diamonds, color)
diamonds_mp_by_color <- summarise(diamonds_by_color, mean_price = mean(price))

diamonds_by_cut <- group_by(diamonds, cut)
diamonds_mp_by_cut <- summarise(diamonds_by_cut, mean_price = mean(price))

library(gridExtra)
clarity <- ggplot(diamonds_mp_by_clarity, aes(x=clarity, y=mean_price)) +
  geom_bar(stat="identity")
color <- ggplot(diamonds_mp_by_color, aes(x=color, y=mean_price)) +
  geom_bar(stat="identity")
cut <- ggplot(diamonds_mp_by_cut, aes(x=cut, y=mean_price)) +
  geom_bar(stat="identity")

grid.arrange(clarity, color, cut)



patents <- read.csv('patents.csv')
names(patents)
colnames(patents) <- c("country", "1984", "1985", "1986", "1987", "1988", "1989", "1990", "1991", "1992", "1993", "1994", "1995", "1996", "1997", "1998", "1999", "2000", "2001", "2002", "X.1")

keep <- c("country", "1984", "1985", "1986", "1987", "1988", "1989", "1990", "1991", "1992", "1993", "1994", "1995", "1996", "1997", "1998", "1999", "2000", "2001", "2002")

patents <- patents[keep]

patents <- na.omit(patents)

#install.packages('tidyr')
library(tidyr)

patents_gathered <- gather(patents, 'year', 'n', 2:20)

summary(patents_gathered$n)

by(patents_gathered$n, patents_gathered$country, summary)

qplot(x = year, y = n, data = patents_gathered)

library(dplyr)

patents_max <- patents_gathered %>% 
  group_by(country) %>%
  filter(n == max(n))
  
ggplot(patents_gathered, aes(x=year,y=n)) +
  geom_point()

ggplot(patents_max, aes(x=country,y=n)) +
  geom_point()

```

Click **KnitHTML** to see all of your hard work and to have an html
page of this lesson, your answers, and your notes!

