Lesson 3
========================================================

***

### What to Do First?
Notes:

***

### Pseudo-Facebook User Data
Notes:

```{r Pseudo-Facebook User Data}
setwd('~/r/nanodegree/explor_data_analys/')

list.files()

pf <- read.csv('OLD/pseudo_facebook.tsv', sep="\t")
```

***

### Histogram of Users' Birthdays
Notes:

```{r Histogram of Users\' Birthdays}
#install.packages('ggplot2')
library(ggplot2)

qplot(x = dob_day, data = pf) +
  scale_x_discrete(breaks=1:31) + 
  facet_wrap(~dob_month, ncol = 3)

qplot(x = friend_count, data = pf)
```

***

#### What are some things that you notice about this histogram?
Response:

***

### Moira's Investigation
Notes:

***

### Estimating Your Audience Size
Notes:

***

#### Think about a time when you posted a specific message or shared a photo on Facebook. What was it?
Response:

#### How many of your friends do you think saw that post?
Response:

#### Think about what percent of your friends on Facebook see any posts or comments that you make in a month. What percent do you think that is?
Response:

***

### Perceived Audience Size
Notes:

***
### Faceting
Notes:

```{r Faceting}

```

#### Let’s take another look at our plot. What stands out to you here?
Response:

***

### Be Skeptical - Outliers and Anomalies
Notes:

***

### Moira's Outlier
Notes:
#### Which case do you think applies to Moira’s outlier?
Response:

***

### Friend Count
Notes:

#### What code would you enter to create a histogram of friend counts?

```{r Friend Count}

```

#### How is this plot similar to Moira's first plot?
Response:

***

### Limiting the Axes
Notes:

```{r Limiting the Axes}

```

### Exploring with Bin Width
Notes:

***

### Adjusting the Bin Width
```{r}
qplot(x = friend_count, data = pf, binwidth = 25) + 
  scale_x_continuous(limits = c(0, 1000), breaks = seq(0, 1000, 50))
```
Notes:

### Faceting Friend Count
```{r Faceting Friend Count}
# What code would you add to create a facet the histogram by gender?
# Add it to the code below.
qplot(x = friend_count, data = pf, binwidth = 10) +
  scale_x_continuous(limits = c(0, 1000),
                     breaks = seq(0, 1000, 50)) +
  facet_wrap(~gender)

qplot(x = friend_count, data = pf, binwidth = 10) +
  scale_x_continuous(limits = c(0, 1000),
                     breaks = seq(0, 1000, 50)) +
  facet_grid(gender ~ .)
```

***

### Omitting NA Values
Notes:

```{r Omitting NA Values}
qplot(x = friend_count, data = subset(pf, !is.na(gender)),
      binwidth = 10) + 
  scale_x_continuous(lim = c(0,1000), breaks = seq(0, 1000, 50)) +
  facet_wrap(~gender)
```

***

### Statistics 'by' Gender
Notes:

```{r Statistics \'by\' Gender}
table(pf$gender)
by(pf$friend_count, pf$gender, summary)
```

#### Who on average has more friends: men or women?
Response:

#### What's the difference between the median friend count for women and men?
Response:

#### Why would the median be a better measure than the mean?
Response:

***

### Tenure
Notes:

```{r Tenure}
qplot(x = tenure, data = pf, binwidth = 30,
      color = I('black'), fill = I('#099DD9'))
```

***

#### How would you create a histogram of tenure by year?

```{r Tenure Histogram by Year}
qplot(x = tenure/365, data = pf, binwidth = 0.5,
      color = I('black'), fill = I('#099DD9')) +
  scale_x_continuous(breaks = seq(1,7,1), limits = c(0,7))
```

***

### Labeling Plots
Notes:

```{r Labeling Plots}
qplot(x = tenure/365, data = pf,
      xlab = 'Number of years using Facebook',
      ylab = 'NUmber of users in sample',
      color = I('black'), fill = I('#F79420')) +
  scale_x_continuous(breaks = seq(1,7,1), limits = c(0,7))
```

***

### User Ages
Notes:
```{r}
summary(pf$age)
```

```{r User Ages}
qplot(x = age, data = pf, binwidth = 1,
      color = I('black'), fill = I('#5760AB')) +
  scale_x_continuous(breaks = seq(0,113,5), limits = c(0,113))
```

#### What do you notice?
Response:

***

### The Spread of Memes
Notes:

***

### Lada's Money Bag Meme
Notes:

***

### Transforming Data
```{r}
qplot(x = friend_count, data = pf)

summary(pf$friend_count)

summary(log10(pf$friend_count + 1))

summary(sqrt(pf$friend_count))
```
Notes:

***

### Add a Scaling Layer
Notes:

```{r Add a Scaling Layer}
#install.packages('gridExtra')
library(gridExtra)

q1 <- qplot(x = friend_count, data = pf)
q2 <- qplot(x = log10(friend_count + 1), data = pf)
q3 <- qplot(x = sqrt(friend_count), data = pf)
grid.arrange(q1, q2, q3, ncol=1)

p1 <- ggplot(aes(x = friend_count), data = pf) + geom_histogram()
p2 <- p1 + scale_x_log10()
p3 <- p1 + scale_x_sqrt()

grid.arrange(p1, p2, p3, ncol = 1)

logScale <- qplot(x = log10(friend_count + 1), data = pf)
countScale <- ggplot(aes(x = friend_count), data = pf) + 
  geom_histogram() +
  scale_x_log10()

grid.arrange(logScale, countScale,  ncol = 2)
```

***


### Frequency Polygons

```{r Frequency Polygons}
qplot(x = friend_count, data = subset(pf, !is.na(gender)),
      binwidth = 10) +
  scale_x_continuous(lim = c(0, 1000), breaks = seq(0, 1000, 50)) +
  facet_wrap(~gender)

qplot(x = friend_count, y = ..count../sum(..count..),
      data = subset(pf, !is.na(gender)),
      xlab = 'Friend Count',
      ylab = 'Proportion of Users with that friend count',
      binwidth = 10, geom = 'freqpoly', color = gender) +
  scale_x_continuous(lim = c(0, 1000), breaks = seq(0, 1000, 50))

ggplot(aes(x = friend_count, y = ..count../sum(..count..)), data = subset(pf, !is.na(gender))) +
  geom_freqpoly(aes(color = gender), binwidth=10) + 
  scale_x_continuous(limits = c(0, 1000), breaks = seq(0, 1000, 50)) +
  xlab('Friend Count') + 
  ylab('Percentage of users with that friend count')
```

***

### Likes on the Web
Notes:

```{r Likes on the Web}
qplot(x = www_likes, data = subset(pf, !is.na(gender)),
      geom = 'freqpoly', color = gender) +
  scale_x_continuous() +
  scale_x_log10()

by(pf$www_likes, pf$gender, sum)
```


***

### Box Plots
Notes:

```{r Box Plots}
qplot(x = gender, y = friend_count,
      data = subset(pf, !is.na(gender)), 
      geom = 'boxplot') +
  coord_cartesian(ylim = c(0, 1000))

by(pf$friend_count, pf$gender, summary)

names(pf)

by(pf$friendships_initiated, pf$gender, mean)
```

#### Adjust the code to focus on users who have friend counts between 0 and 1000.

```{r}

```

***

### Box Plots, Quartiles, and Friendships
Notes:

```{r Box Plots, Quartiles, and Friendships}

```

#### On average, who initiated more friendships in our sample: men or women?
Response:
#### Write about some ways that you can verify your answer.
Response:
```{r Friend Requests by Gender}

```

Response:

***

### Getting Logical
Notes:

```{r Getting Logical}
summary(pf$mobile_likes)

summary(pf$mobile_likes > 0)

pf$mobile_check_in <- NA
pf$mobile_check_in <- ifelse(pf$mobile_likes > 0, 1, 0)
pf$mobile_check_in <- factor(pf$mobile_check_in)
summary(pf$mobile_check_in)

sum(pf$mobile_check_in == 1)/length(pf$mobile_check_in)
```

```{r}
patents <- read.csv('patents.csv')
names(patents)
colnames(patents) <- c("country", "1984", "1985", "1986", "1987", "1988", "1989", "1990", "1991", "1992", "1993", "1994", "1995", "1996", "1997", "1998", "1999", "2000", "2001", "2002", "X.1")

keep <- c("country", "1984", "1985", "1986", "1987", "1988", "1989", "1990", "1991", "1992", "1993", "1994", "1995", "1996", "1997", "1998", "1999", "2000", "2001", "2002")

patents <- patents[keep]

patents <- na.omit(patents)

install.packages('tidyr')
library(tidyr)

patents_gathered <- gather(patents, 'year', 'n', 2:20)

summary(patents_gathered$n)

by(patents_gathered$n, patents_gathered$country, summary)

qplot(x = year, y = n, data = patents_gathered)

install.packages('dplyr')
library(dplyr)

patents_max <- patents_gathered %>% 
  group_by(country) %>%
  filter(n == max(n))
  
ggplot(patents_max, aes(x=country,y=n)) +
  geom_point()
```

```{r}
bdays <- read.csv('birthdaysExample.csv')

bdays$dates <- strptime(bdays$dates, format="%D")

sum(is.na(bdays$dates))

bdays_nona <- bdays[!is.na(bdays$dates),]

years <- strftime(bdays_nona, format="%y")
months_days <- strftime(bdays_nona, format="%m-%d")
months <- strftime(bdays_nona, format="%m")
days <- strftime(bdays_nona, format="%d")

qplot(months)

```

Response:

***

### Analyzing One Variable
Reflection:

***

Click **KnitHTML** to see all of your hard work and to have an html
page of this lesson, your answers, and your notes!