Which chemical properties influence the quality of red wines?
========================================================
###Austin J. Alexander

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using
# in your analysis in this code chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk.
# This prevents the code from displaying in the knitted HTML output.
# You should set echo=FALSE for all code chunks in your file.

library(ggplot2)
library(reshape2)
library(GGally)
library(grid)
library(gridExtra)
library(dplyr)
library(memisc)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
setwd('~/r/nanodegree/explor_data_analys/')
#list.files()
reds <- read.csv('wineQualityReds.csv')
```

------

## Check out the data
##### show column names
```{r echo=FALSE}
names(reds[,2:13])
```

##### basic stats for each column
```{r echo=FALSE}
summary(reds[,2:13])
```

##### check for any NaN values
```{r echo=FALSE}
sum(is.na(reds))
```

##### check the distribution of quality
```{r echo=FALSE}
table(reds$quality)
```

As can be seen above, the vast majority of wines are scored 5 or 6.
```{r echo=FALSE}
# break up the quality scores into quality buckets of 2 scores each
reds_3_4 <- subset(reds, quality == 3 | quality == 4)
reds_5_6 <- subset(reds, quality == 5 | quality == 6)
reds_7_8 <- subset(reds, quality == 7 | quality == 8)
# take bottom 2 and top 2
reds_3_4_7_8 <- merge(reds_3_4, reds_7_8, all =TRUE)
# take bottom and top
reds_3_8 <- subset(reds, quality == 3 | quality == 8)
```

##### proportions 
3's and 4's
```{r echo=FALSE}
sum(!is.na(reds_3_4))/sum(!is.na(reds))
```

5's and 6's
```{r echo=FALSE}
sum(!is.na(reds_5_6))/sum(!is.na(reds))
```

7's and 8's
```{r echo=FALSE}
sum(!is.na(reds_7_8))/sum(!is.na(reds))
```

Scores of 5 and 6 account for over 82% percent of all scores!
This suggests that the most useful information might be found by examining the lowest and highest scorers, but we'll save that for later.

## Univariate Plots Section
##### visualize quality
```{r echo=FALSE}
ggplot(reds, aes(x=factor(quality))) +
  geom_bar(aes(colour = factor(quality)))
```

Examine correlations (technically these are bivariate plots, but the correlation coefficients are displayed nicely).
```{r echo=FALSE}
# subset reds (arbitrarily) by index number for use in ggpairs
ggpairs(reds[c(2,3,4,13)])
ggpairs(reds[c(5,6,7,8,13)])
ggpairs(reds[c(9,10,11,12,13)])
```

Alcohol, volatile.acidity, and sulphates present correlation coefficient values furthest from zero, so let's examine these further.

##### facet alcohol by quality
```{r echo=FALSE}
ggplot(reds, aes(x = alcohol)) +
  geom_histogram() +
  facet_grid(quality ~ .)
```

##### view alcohol as a histogram colored by quality 
```{r echo=FALSE}
ggplot(reds, aes(x = alcohol, fill = as.factor(quality))) +
  geom_histogram()
```

##### facet volatile.acidity by quality
```{r echo=FALSE}
ggplot(reds, aes(x = volatile.acidity)) +
  geom_histogram() +
  facet_grid(quality ~ .)
```

##### view volatile.acidity as a histogram colored by quality 
```{r echo=FALSE}
ggplot(reds, aes(x = volatile.acidity, fill = as.factor(quality))) +
  geom_histogram()
```

##### facet sulphates by quality
```{r echo=FALSE}
ggplot(reds, aes(x = sulphates)) +
  geom_histogram() +
  facet_grid(quality ~ .)
```

##### view sulphates as a histogram colored by quality 
```{r echo=FALSE}
ggplot(reds, aes(x = sulphates, fill = as.factor(quality))) +
  geom_histogram()
```

# Univariate Analysis

### What is the structure of your dataset?
There are 1,599 observations of red wines with 12 recorded features for each observation. Some of the features are related to each other (e.g., those related to acidity). Quality is the only categorical feature.

### What is/are the main feature(s) of interest in your dataset?
Alcohol, volatile.acidity, and sulphates seem to be the features most highly correlated with quality scores.

### What other features in the dataset do you think will help support your investigation into your feature(s) of interest?
It was unclear what features will be useful at this point.

### Did you create any new variables from existing variables in the dataset?
No, there didn't seem to be much of a need to create new variables.

### Of the features you investigated, were there any unusual distributions? Did you perform any operations on the data to tidy, adjust, or change the form of the data? If so, why did you do this?
None of the features seemed unusual enough to explore futher and, no, I didn't bother tidying/adjusting the form of the data at this point.

------

## Bivariate Plots Section
Continue to focus on alcohol, volatile.acidity, and sulphates.

##### alcohol vs (jittered) quality colored by quality
```{r echo=FALSE}
ggplot(reds, aes(x = alcohol, y = factor(quality))) +
  geom_jitter(aes(colour = factor(quality)))
```

##### volatile.acidity vs (jittered) quality colored by quality
```{r echo=FALSE}
ggplot(reds, aes(x = volatile.acidity, y = factor(quality))) +
  geom_jitter(aes(colour = factor(quality)))
```
  
##### sulphates vs (jittered) quality colored by quality
```{r echo=FALSE}
ggplot(reds, aes(x = sulphates, y = factor(quality))) +
  geom_jitter(aes(colour = factor(quality)))
```

##### boxplot of alcohol colored by quality
```{r echo=FALSE}
ggplot(reds, aes(factor(quality), alcohol, fill=factor(quality))) +
  geom_boxplot()
```

##### boxplot of volatile.acidity colored by quality
```{r echo=FALSE}
ggplot(reds, aes(factor(quality), volatile.acidity, fill=factor(quality))) +
  geom_boxplot()
```

##### boxplot of sulphates colored by quality
```{r echo=FALSE}
ggplot(reds, aes(factor(quality), sulphates, fill=factor(quality))) +
  geom_boxplot()
```

##### alcohol vs volatile.acidity faceted by quality
```{r echo=FALSE}
ggplot(reds, aes(x = alcohol, y = volatile.acidity)) +
  geom_point(aes(color = factor(quality))) +
  facet_grid(. ~ quality)
```

##### alcohol vs sulphates faceted by quality
```{r echo=FALSE}
ggplot(reds, aes(x = alcohol, y = sulphates)) +
  geom_point(aes(color = factor(quality))) +
  facet_grid(. ~ quality)
```

##### group reds by quality
```{r}
quality_groups <- group_by(reds, quality)
grouped_reds <- summarise(quality_groups, 
                alcohol_mean = mean(alcohol),
                volatile_acid_mean = mean(volatile.acidity),
                sulphates_mean = mean(sulphates),
                n = n())
grouped_reds <- arrange(grouped_reds, quality)
```

##### quality vs alcohol mean
```{r echo=FALSE}
ggplot(grouped_reds, aes(x = quality, y = alcohol_mean)) + 
  geom_line()
```

##### quality vs alcohol with quantile lines
```{r echo=FALSE}
ggplot(reds, aes(x = quality, y = alcohol)) +
  geom_point(alpha = 0.5,
             position = position_jitter(h = 0),
             color = 'orange') +
  coord_trans(y = 'sqrt') +
  geom_line(stat = 'summary', fun.y = quantile, probs = 0.1,
            linetype = 2, color = 'blue') +
  geom_line(stat = 'summary', fun.y = quantile, probs = 0.5,
            color = 'blue') +
  geom_line(stat = 'summary', fun.y = quantile, probs = 0.9,
            linetype = 2, color = 'blue')
```

##### quality vs volatile.acidity mean
```{r echo=FALSE}
ggplot(grouped_reds, aes(x = quality, y = volatile_acid_mean)) + 
  geom_line()
```

##### quality vs volatile.acidity with quantile lines
```{r echo=FALSE}
ggplot(reds, aes(x = quality, y = volatile.acidity)) +
  geom_point(alpha = 0.5,
             position = position_jitter(h = 0),
             color = 'orange') +
  coord_trans(y = 'sqrt') +
  geom_line(stat = 'summary', fun.y = quantile, probs = 0.1,
            linetype = 2, color = 'blue') +
  geom_line(stat = 'summary', fun.y = quantile, probs = 0.5,
            color = 'blue') +
  geom_line(stat = 'summary', fun.y = quantile, probs = 0.9,
            linetype = 2, color = 'blue')
```

##### quality vs sulphates mean
```{r echo=FALSE}
ggplot(grouped_reds, aes(x = quality, y = sulphates_mean)) + 
  geom_line()
```

##### quality vs sulphates with quantile lines
```{r echo=FALSE}
ggplot(reds, aes(x = quality, y = sulphates)) +
  geom_point(alpha = 0.5,
             position = position_jitter(h = 0),
             color = 'orange') +
  coord_trans(y = 'sqrt') +
  geom_line(stat = 'summary', fun.y = quantile, probs = 0.1,
            linetype = 2, color = 'blue') +
  geom_line(stat = 'summary', fun.y = quantile, probs = 0.5,
            color = 'blue') +
  geom_line(stat = 'summary', fun.y = quantile, probs = 0.9,
            linetype = 2, color = 'blue')
```

##### alcohol vs volatile.acidity colored by quality
```{r echo=FALSE}
ggplot(reds, aes(x = alcohol, y = volatile.acidity)) +
  geom_point(aes(color = factor(quality)))
```

##### with quality smoothing lines
```{r echo=FALSE}
ggplot(reds, aes(x = alcohol, y = volatile.acidity, colour = factor(quality))) +
  stat_smooth(method = lm, aes(fill = factor(quality))) +
  geom_point()
```

##### now only 3's, 4's, 7's, and 8's
```{r echo=FALSE}
ggplot(reds_3_4_7_8, aes(x = alcohol, y = volatile.acidity, colour = factor(quality))) +
  geom_point()
```

##### with quality smoothing lines
```{r echo=FALSE}
ggplot(reds_3_4_7_8, aes(x = alcohol, y = volatile.acidity, colour = factor(quality))) +
  stat_smooth(method = lm, aes(fill = factor(quality))) +
  geom_point()
```

##### now only 3's and 8's
```{r echo=FALSE}
ggplot(reds_3_8, aes(x = alcohol, y = volatile.acidity, colour = factor(quality))) +
  stat_smooth(method = lm, aes(fill = factor(quality))) +
  geom_point()
```

##### alcohol vs sulphates
```{r echo=FALSE}
ggplot(reds_3_4_7_8, aes(x = alcohol, y = sulphates)) +
  geom_point(aes(color = factor(quality)))
```

##### with quality smoothing lines
```{r echo=FALSE}
ggplot(reds_3_4_7_8, aes(x = alcohol, y = sulphates, colour = factor(quality))) +
  stat_smooth(method = lm, aes(fill = factor(quality))) +
  geom_point()
```

##### now only 3's and 8's
```{r echo=FALSE}
ggplot(reds_3_8, aes(x = alcohol, y = sulphates, colour = factor(quality))) +
  stat_smooth(method = lm, aes(fill = factor(quality))) +
  geom_point()
```

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. How did the feature(s) of interest vary with other features in the dataset?
Simply put: alcohol, volatile.acidity, and sulphates (particularly the first two) appear to have an affect of the quality scores. Alcohol will be discussed below, but, in general, the lower the volatile acidity, the higher the quality score; the inverse is true for sulphates and quality score.

### Did you observe any interesting relationships between the other features (not the main feature(s) of interest)?
I didn't bother looking at the other features because I'm focusing on answering the primary question driving this project.

### What was the strongest relationship you found?
Alcohol. Funny enough, a higher alcohol content seems to encourage a higher score.

------

# Multivariate Plots Section

##### alcohol vs volatile.acidity, colored by alcohol, faceted by quality
```{r echo=FALSE}
ggplot(reds, aes(x = alcohol, y = volatile.acidity)) +
  geom_point(aes(colour = alcohol), alpha=0.7) +
  scale_colour_gradient(low = 'blue', high = 'orange') +
  facet_grid(. ~ quality)
```

##### alcohol vs sulphates, colored by alcohol, faceted by quality
```{r echo=FALSE}
ggplot(reds, aes(x = alcohol, y = sulphates)) +
  geom_point(aes(colour = alcohol), alpha=0.7) +
  scale_colour_gradient(low = 'purple', high = 'orange') +
  facet_grid(. ~ quality)
```

##### alcohol vs volatile.acidity, colored by quality, sized by sulphates
```{r echo=FALSE}
ggplot(reds, aes(x = alcohol, y = volatile.acidity)) + 
  geom_point(aes(colour = factor(quality), size = sulphates), alpha = 0.7)
```

##### alcohol vs volatile.acidity, colored by quality, sized by sulphates, with contour lines showing overall quantity clustering
```{r echo=FALSE}
ggplot(reds, aes(x = alcohol, y = volatile.acidity)) + 
  geom_point(aes(colour = factor(quality), size = sulphates), alpha = 0.7) +
  geom_density2d(show_guide=FALSE)
```

##### alcohol vs volatile.acidity with only contour lines colored by quality showing quality clusters
```{r echo=FALSE}
ggplot(reds, aes(x = alcohol, y = volatile.acidity)) + 
  geom_density2d(aes(colour = factor(quality)))
```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. Were there features that strengthened each other in terms of looking at your feature(s) of interest?
As my examination continued, I felt better and better about the apparent relationship between alcohol, volatile.acidity, and sulphates and quality scores.

### Were there any interesting or surprising interactions between features?
I found it interesting that sulphate levels seem to have a sweet spot when it comes to quality scores.

### OPTIONAL: Did you create any models with your dataset? Discuss the strengths and limitations of your model.

##### attempts at a simple linear regression model
```{r echo=FALSE}
m1 <- lm(quality ~ alcohol, data = reds)
m2 <- update(m1, ~ . + volatile.acidity)
m3 <- update(m2, ~ . + sulphates)
m4 <- lm(quality ~ alcohol + volatile.acidity + sulphates, data = reds)
m5 <- lm(quality ~ alcohol : volatile.acidity : sulphates, data = reds)
m6 <- lm(quality ~ alcohol * volatile.acidity * sulphates, data = reds)

mtable(m1,m2,m3,m4,m5,m6)
```
Since model 6 had the highest R^2 value, I tested it with some obvious extreme cases (based on what seems to have been discovered above) using only values for alcohol, volatile.acidity, and sulphates:

1. A fake _good_ red was created using a high alcohol content, low volatile.acidity, and average (median) sulphates level and is expected to have a high quality score. 

##### fake good red quality score prediction (95% confidence interval)
```{r echo=FALSE}
fake_good_red = data.frame(alcohol = max(reds$alcohol),
                    volatile.acidity = min(reds$volatile.acidity), 
                    sulphates = median(reds$sulphates))
predict(m6, newdata = fake_good_red, interval="prediction", level = .95)
```

2. A fake _mid_ red was created using an average (median) alcohol content, average (median) volatile.acidity, and average (median) sulphates level and is expected to have a mid-range quality score.

##### fake mid-range red quality score prediction (95% confidence interval)
```{r echo=FALSE}
fake_mid_red = data.frame(alcohol = median(reds$alcohol),
                    volatile.acidity = median(reds$volatile.acidity), 
                    sulphates = median(reds$sulphates))
predict(m6, newdata = fake_mid_red, interval="prediction", level = .95)
```

3. A fake _bad_ red was created using a low alcohol content, high volatile.acidity, and low sulphates level and is expected to have a low quality score.

##### fake bad red quality score prediction (95% confidence interval)
```{r echo=FALSE}
fake_bad_red = data.frame(alcohol = min(reds$alcohol),
                    volatile.acidity = max(reds$volatile.acidity), 
                    sulphates = min(reds$sulphates))
predict(m6, newdata = fake_bad_red, interval="prediction", level = .95)
```
As should be somewhat expected from the entire investigation so far, combined with the not-too-shabby R^2 value of the simple linear regression model we selected, these predictions were spot on.   

Strength of this model: it works for the obvious cases. Weakness of this model: it's unclear how robust it is.

------

# Final Plots and Summary

### Plot One
##### alcohol vs volatile.acidity, colored by alcohol, sized by sulphates, faceted by quality
```{r echo=FALSE}
ggplot(reds, aes(x = alcohol, y = volatile.acidity)) +
  geom_point(aes(colour = alcohol, size = sulphates), alpha=0.7) +
  scale_colour_gradient(low = 'blue', high = 'orange') +
  facet_grid(. ~ quality) +
  ggtitle('Alcohol vs Volatile.Acidity by Quality Score')
```

### Description One
This plot makes it easy to see the distribution of quality scores (most are in the middle), the rightward trend of alcohol content, the downward slope of volatile.acidity, and the mid-range sweet-spot of sulphates levels all in relation to quality scoring.

### Plot Two
##### quality vs alcohol, colored by volatile.acidity, sized by sulphates, with a smoothing line
```{r echo=FALSE}
ggplot(reds, aes(x = quality, y = alcohol)) +
  geom_point(alpha = 0.7,
             position = position_jitter(h = 0),
             aes(colour = volatile.acidity,
                 size = sulphates)) +
  scale_colour_gradient(low = 'blue', high = 'yellow') +
  stat_smooth(method = 'lm', colour = 'green') +
  ggtitle('Quality vs Alcohol')
```

### Description Two
Although alcohol and quality are swapped from their perhaps expected axis locations, the swapping, along with the smoothing line, makes it clear that as quality increases, so do alcohol content (and, thus, the reverse relationship is true). Volatile.acidity and sulphates continue to play supporting roles. 

### Plot Three
##### alcohol vs volatile.acidity with only clusters colored by quality
```{r echo=FALSE}
ggplot(reds, aes(x = alcohol, y = volatile.acidity)) + 
  stat_density2d(aes(fill = factor(quality)), geom = 'polygon', alpha = 0.5) +
  ggtitle('Alcohol vs Volatile.Acidity : Quality Clusters')
```

### Description Three
Perhaps my favorite plot, borrowing from the experimentation with contours earlier, this plot, although leaving out sulphates, makes it clear that there are distinct clusters of quality scores that are quite obviously related to volatile.acidity and alcohol levels. If I were given a new red wine with only those two features listed, I would be very confident using merely this plot to predict the quality score (assuming the same wine experts responsible for this data set).

------

# Reflection
A fruitful exercise, this project exposed two or three features of red wines that, when related to one another, seem to lead to obvious groupings. Alcohol, volatile.acidity, and sulphates (in that order) appear to affect the (perceived) quality of red wines, at least among those wine experts consulted in the making of this data set.

